---
title: "Land use effects on Pacific Giant Salamander population and weight"
subtitle: " A report prepared for ESM 206"
author: "Maggie Brickner"
date: "12/04/2019"
output: html_document
---

### Introduction

The Pacific Giant Salamander or Coastal Giant Salamander (*Dicamptodon tenebrosus*) is among the largest terrestrial salamanders in the world. They live in coastal, wet forests usually near water. Growing up to 30 cm (11.8 inches) long, they feed on insects and slugs as well as other amphibians and even small rodents (Naish 2016). Logging is a major threat to the Pacific Giant Salamander. In previous studies logging has been shown to severely decrease the population of Pacific Giant Salamander larvae in streams (AmphibiaWeb 2010). Here I use data collected along Mack Creek within the H.J. Andrews Experimental Forest located in the Cascade Mountain Range in Oregon. Data has been collected on Pacific Giant Salamanders in this area since 1993 in areas that were clearcut in 1963 and in an old growth forest (Gregory 2016). 


![**Figure 1.** Mature Pacific Giant Salamader. Credit: [Scientific American](https://blogs.scientificamerican.com/tetrapod-zoology/the-pacific-giant-salamanders/).](coastal_salamander_photo.jpg) 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
#Attach the needed packages. This is a work in progress so for now I'll just attach the tidyverse and janitor. More to come though!  
library(tidyverse)
library(janitor)
library(kableExtra)
library(ggbeeswarm)
library(car)

#Read in the data and check it out!

salamander <- read_csv("mack_creek_vertebrates.csv") %>%
  janitor::clean_names() %>% 
  filter(species == "DITE")

#it seems like the data is tidy in other ways. I may have to change the data format eventually. This will be my starting point!
```

### Data and Methods

![**Figure 2.** Study location overlaid on the Andrews Forest informational map. Credit: [H.J. Andrews Experimental Forest](https://andrewsforest.oregonstate.edu/sites/default/files/lter/pubs/pdf/pub3654.pdf).](study_location.png)


##### Results

##### A.

```{r}
# We need to first filter by the locations in the data set and then plot it. This is the perfect place to make my ggridges_density dreams come true. Never mind it looked bad

salamander_cc_og <- salamander %>% 
  filter(section %in% c("OG", "CC")) %>% 
  mutate(section_name = ifelse(section == "CC", "Clearcut", "Old Growth"))
  

#Ok I think that was useless and already done...nevermid I renamed the sections for nicer graphing!

ggplot(data = salamander_cc_og, aes(x = year)) +
  geom_histogram(binwidth = 1, aes(fill = section_name)) +
  theme_bw() + 
  scale_x_continuous(expand = c(0,0), breaks = seq(1993,2017, by = 3)) +
  scale_y_continuous(expand = c(0,1)) +
  labs(title = "Pacific giant salamander observations", subtitle = "1993-2017", x = "Year", y = "Counts")+
  facet_wrap(~section_name) +
  theme(panel.spacing = unit(2, "lines"), 
        plot.title = element_text(hjust = 0.5), 
        plot.subtitle = element_text(hjust = 0.5), 
        legend.position = "none")
```

##### B. 

```{r}
# We arent interested in salamanders in the side pool, so I need to filter out the pools. I think I'll also need to eventually rename for easier graphing. 

salamander_channel <- salamander %>% 
  mutate(section_name = ifelse(section == "CC", "Clearcut", "Old Growth")) %>% 
  mutate(unit_name = ifelse(unittype == "C", "Cascade", ifelse(unittype == "P", "Pool", "Side Channel"))) %>% 
  filter(unit_name %in% c("Cascade", "Pool", "Side Channel")) %>% 
  filter(year == 2017) %>% 
  count(unit_name, section_name)

# Now we want to create a table of the counts and proportions

salamander_channel_table <- salamander_channel %>% 
  pivot_wider(names_from = unit_name, values_from = n)

# It looks like it worked! ok, now time to make it look nice
channel_unit <- salamander_channel_table %>% 
  janitor::adorn_percentages(denominator = "row") %>% 
  adorn_pct_formatting(digits = 1) %>% 
  adorn_ns(position = "front")

# Ooooh pretty! Now to the kable

kable(channel_unit, col.names = c("Section", "Cascade", "Pool", "Side Channel")) %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>% 
  add_header_above(c("2017 Salamander Location" = 4))


```

##### C.

H~0~: There is no significant difference in habitat type population between clear cut and old growth forests
H~A~: There is a significant difference in habitat population between clear cut and old growth forests

```{r}

# We're comparing counts so this will be a chi-square test. Now some preparation

salamander_chi <- salamander_channel_table %>% 
  select (-section_name)

# And now for the test

salamander_chi <- chisq.test(salamander_chi)

salamander_chi

```

There is not sufficient evidence to reject the null hypothesis ($\chi^2$(`r salamander_chi$parameter`) = `r round(salamander_chi$statistic, 2)`, *p* = `r round(salamander_chi$p.value,2)`)

*how to get n values in this satatement. I guess you could always just type*

##### D.
H~0~: There is no significant difference in mean salamander weights between clear cut and old growth forests
H~A~: There is a significant difference in mean salamander weights between clear cut and old growth forests
```{r}
# Since I've done some funky stuff with the above information I'm going to start fresh with the data frame

salamander_wt <- salamander %>% 
  filter(year == 2017, section %in% c("CC", "OG"))

# first we should look at the data to see if it is normally distributed
ggplot(data = salamander_wt, aes(x = weight)) +
  geom_density() + 
  facet_wrap(~section)

ggplot(data = salamander_wt, aes(sample = weight)) +
  geom_qq() +
  facet_wrap(~section)

# oooh looks like not normally distributed, but we still have a large enough sample size to compare per the central limit theorom. Now time to do a two-sample t-test
salamander_wt_cc <- salamander_wt %>% 
  filter(section == "CC")

salamander_wt_og <- salamander_wt %>% 
  filter(section == "OG")

salamander_wt_ttest <- t.test(salamander_wt_cc$weight, salamander_wt_og$weight)

salamander_wt_ttest

# We cannot reject the null hypothesis. 

```

##### E.

```{r}
# I want to create a beeswarm plot. First to sort out the data as I want to. I also want to drop the salamander that did not have an observed weight. It gives me errors for all C habitats

salamander_wt_habitat <- salamander %>% 
  filter(year == 2017, section %in% c("CC", "OG"), unittype %in% c("C", "P", "SC")) %>% 
  drop_na(weight) %>% 
  group_by(unittype)

# Next I need to create some summary info so I can pull it into the beeswarm

salamander_wt_summary <- salamander_wt_habitat %>% 
  summarize(mean_salamander_wt = mean(weight), 
            sd_salamander_wt = sd(weight), 
            sample_size = n(), 
            se_salamander_wt = sd(weight)/ sqrt(n()),
            var_salamander_wt = var(weight))
  
# Next to create the graph

ggplot() +
  geom_beeswarm(data = salamander_wt_habitat, aes(x = unittype, y = weight),
                size = 1, 
                alpha = 0.6, 
                color = "gray60") +
  theme_bw() +
  scale_y_continuous(expand = c(0,1)) +
  scale_x_discrete(labels = c("Cascades", "Pools", "Side-channels")) +
  geom_point(data = salamander_wt_summary, 
             aes(x = unittype, y = mean_salamander_wt), 
             color = "red", 
             size = 1.5) +
  geom_errorbar(data = salamander_wt_summary, 
                aes(x = unittype,
                    ymin = mean_salamander_wt - se_salamander_wt, 
                    ymax = mean_salamander_wt + se_salamander_wt),
                color = "red",
                width = 0.1) +
  labs(title = "Salamander weights and habitat", subtitle = "2017", x = "Channel Habitat", y = "Weight (g)")


# the error I'm getting is fine since there is one NA value for weight. Nevermind! I got rid of that as it was causing more problems. Should I change the minimums since they're 

```

```{r}
# Anova test (comparing multiple means) First we want to see if variances are the same, because that's what we assume when using an ANOVA

leveneTest(weight ~ unittype, data = salamander_wt_habitat)

# with an alpha of 0.05 we do not have sufficient evidence to reject the null hypothesis that the means are the same. 

```
```{r}
# great! Now we can do an ANOVA!

salamander_aov <- aov(weight ~ unittype, data = salamander_wt_habitat)

summary(salamander_aov)

# Oooh finaly some differentiaton we can reject the null hypothesis! Now for a post hoc test

TukeyHSD(salamander_aov)

# From Tukey it looks like the only distinction is between side channel and pool living salamanders in that Pool salamaders are heavier than side channel salamanders. 

```

##### References

AmphibiaWeb. 2010. Dicamptodon tenebrosus: Pacific Giant Salamander. University of California, Berkeley. Retrieved from http://amphibiaweb.org/species/3867


Gregory S. V. 2016. Aquatic Vertebrate Population Study in Mack Creek, Andrews Experimental Forest, 1987 to present. Environmental Data Initiative. Retrieved from https://doi.org/10.6073/pasta/5de64af9c11579266ef20da2ff32f702 

Naish, D. 2016. The Pacific Giant Salamanders. Scientific American Tetrapod Zoology Blog. Retrieved from https://blogs.scientificamerican.com/tetrapod-zoology/the-pacific-giant-salamanders/